Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-13 article.tex
rework article to place focus on error-reporting using transformers drop example showing whitespace/comment logging as it's not as important as the error reporting

\title{Combinator Parsers:  a Monad Transformer Approach}
\author{Matt Fenwick\email{mfenwick100@gmail.com}}

@@

\begin{introduction}
Parsing, or "the process of analysing a 
string of symbols according to the rules of a formal grammar" (Wikipedia),
is one of the most fundamental and interesting problems in programming.
The general goal of parsing is to build a tree representing the structure
of the parsed input; further operations such as interpretation or
code generations are then performed on the tree.

However, in addition to recognizing a stream of tokens and building a parse
tree, a practical parser has many additional duties.  If parsing fails,
the location and cause of the failure must be accurately reported so that
the user can address the problem; specially formatted comments may be need
to be capture in order to be used
to build hyperlinked code documentation, instead of simply thrown away; 
failed parses may be accompanied by partial results to save the cost
of reparsing the input before the error.

Monad transformers offer an elegant solution, both to the problem of 
implementing parser combinators themselves, as well as to the issue of
building clean and modular parsers that support multiple effects.
We'll explore what they have to offer using a simple language -- Woof --
and build a number of separate parsers for it!
\end{introduction}

@@

\section{Example 1: Recognition and Tree-Building}
The first parser we build will be responsible for determining whether input
text conforms to the language definition and simultaneously build an 
Abstract Syntax Tree representing of the structure of the recognized input.

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
Right' (Just ([ADefine "fgh" 
                       (ALambda ["f","x","y"] 
                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
                                 AApp (ASymbol "f") [ASymbol "y",
                                                     ASymbol "a",
                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
              []))
\end{verbatim}
Here are examples showing errors detected by the parser:
\begin{verbatim}
ghci$ map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                     "{define a}", "{define a b", 
                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
Left' ("application: missing operator",('(',1,1))
Left' ("application: missing close parenthesis",('(',1,1))
Left' ("define: missing symbol",('{',1,1))
Left' ("define: missing form",('{',1,1))
Left' ("define: missing close curly",('{',1,1))
Left' ("lambda: missing parameter list",('{',1,1))
Left' ("lambda: duplicate parameter names",('{',1,1))
Left' ("lambda: missing parameter list close curly",('{',1,1))
Left' ("lambda: missing body form",('{',1,1))
Left' ("lambda: missing close curly",('{',1,1))
Left' ("special form: unable to parse",('{',1,1))
Left' ("woof: unparsed input",(',',1,2))
\end{verbatim}

\section{Example 3: whitespace/comment-logging}

@@

\section{Limitations of the Monad Transformer Approach}

\section{Conclusion}

Parser combinators and monad transformers FTW!

>>>>

\title{Error Reporting Parsers:  a Monad Transformer Approach}
\author{Matt Fenwick\email{mfenwick100@gmail.com}}

@@

\begin{introduction}
Parsing has long been one of the most fascinating and challenging applications
of programming, as evidenced by the large number of parsing libraries available
today.  However, in addition to recognizing a stream of tokens and building a parse
tree, a practical parser has many additional duties.  If parsing fails,
the location and cause of the failure must be accurately reported so that
the user can address the problem
Thus, building composable parsers which accurately report any errors
is extremely difficult.  In this article, we'll see how building
a parsing library around monad transformers allows to build clean and simple 
parsers which nevertheless provide relevant and location-specific error messages.
At the same time, we'll explore the design and some of the tradeoffs of the 
popular Haskell library for monad transformers.
\end{introduction}

@@

\section{Example 1: Recognition and Tree-Building}

Parsing is "the process of analysing a 
string of symbols according to the rules of a formal grammar" (Wikipedia)
The general goal of parsing is to build a tree representing the structure
of the parsed input; further operations such as interpretation or
code generations are then performed on the tree.

The first parser we build will be responsible for determining whether input
text conforms to the language definition and simultaneously build an 
Abstract Syntax Tree representing of the structure of the recognized input.

@@

\subsection{Error messages:  examples}
While correct input will be parsed in the same way by our new parsers, incorrect input will now (hopefully) be
accurately reported as such.  Successful results will be slightly different, however -- there is now an extra
constructor wrapper around the value:
\begin{verbatim}
runParser woof (countLineCol "{define fgh {lambda {f x y} {define a (b c)} (f y a (a x))}}")
Right' (Just ([ADefine "fgh" 
                       (ALambda ["f","x","y"] 
                                [ADefine "a" (AApp (ASymbol "b") [ASymbol "c"]),
                                 AApp (ASymbol "f") [ASymbol "y",
                                                     ASymbol "a",
                                                     AApp (ASymbol "a") [ASymbol "x"]]])],
              []))
\end{verbatim}
Here are examples showing errors detected by the parser:
\begin{verbatim}
ghci$ map (runParser woof . countLineCol) ["()", "(a b", "{define (a b c)}",  
                                     "{define a}", "{define a b", 
                                     "{lambda (a b c)}", "{lambda {a b a} (c d)}", 
                                     "{lambda {a b (c d)}", "{lambda {a b}}", 
                                     "{lambda {a b} (c d)", "{defin x y}", "a,b"]
Left ("application: missing operator",('(',1,1))
Left ("application: missing close parenthesis",('(',1,1))
Left ("define: missing symbol",('{',1,1))
Left ("define: missing form",('{',1,1))
Left ("define: missing close curly",('{',1,1))
Left ("lambda: missing parameter list",('{',1,1))
Left ("lambda: duplicate parameter names",('{',1,1))
Left ("lambda: missing parameter list close curly",('{',1,1))
Left ("lambda: missing body form",('{',1,1))
Left ("lambda: missing close curly",('{',1,1))
Left ("special form: unable to parse",('{',1,1))
Left ("woof: unparsed input",(',',1,2))
\end{verbatim}

\section{Monad Transformer Stacks}

@@

\section{Conclusion}
Furthur applications of monad transformers to parsing:  junk logging, partial results, ???
